---
title: "ASURAT"
author: "Keita Iida"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_section: true
    fig_width: 6.2
    fig_height: 5.2
vignette: >
  %\VignetteIndexEntry{ASURAT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Installations
Attach necessary libraries:
```{r, message = FALSE, warning = FALSE}
library(ASURAT)
library(SingleCellExperiment)
library(SummarizedExperiment)
```



<br>

# Goal
A goal of ASURAT is to cluster and characterize individual samples (cells)
in terms of cell type (or disease), biological function, signaling pathway
activity, and so on (see [here](#multilayered)).



<br>

# Quick start by SingleCellExperiment objects
Having a SingleCellExperiment object (e.g., `sce`), one can use ASURAT by
confirming the following requirements:

* `assays(sce)` contains gene expression data with row and column names as
  variable (gene) and sample (cell), respectively,

If `sce` contains normalized expression data (e.g., `assay(sce, "logcounts")`),
set `assay(sce, "centered")` by subtracting the data with the mean expression
levels across samples (cells).
```{r, eval = FALSE}
mat <- as.matrix(assay(sce, "logcounts"))
assay(sce, "centered") <- sweep(mat, 1, apply(mat, 1, mean), FUN = "-")
```

One may use a Seurat function `Seurat::as.SingleCellExperiment()` for converting
Seurat objects into SingleCellExperiment ones.

Now, ready for the next step [here](#sign).



<br>

# Preprocessing
## Prepare SingleCellExperiment objects
Load single-cell RNA sequencing (scRNA-seq) data.
```{r, message = FALSE}
sce <- TENxPBMCData::TENxPBMCData(dataset = c("pbmc4k"))
pbmc_counts <- as.matrix(assay(sce, "counts"))
rownames(pbmc_counts) <- make.unique(rowData(sce)$Symbol_TENx)
colnames(pbmc_counts) <- make.unique(colData(sce)$Barcode)
```
Here, `pbmc_counts` is a read count table of peripheral blood mononuclear cells
(PBMCs).

Below is a head of `pbmc_counts`:
```{r}
pbmc_counts[1:5, 1:3]
```

Create SingleCellExperiment objects by inputting gene expression data.
```{r}
pbmc <- SingleCellExperiment(assays = list(counts = pbmc_counts),
                             rowData = data.frame(gene = rownames(pbmc_counts)),
                             colData = data.frame(cell = colnames(pbmc_counts)))
```

Check data sizes.
```{r}
dim(pbmc)
```



<br>

## Control data quality
Remove variables (genes) and samples (cells) with low quality,
by processing the following three steps:

1. remove variables based on expression profiles across samples,
2. remove samples based on the numbers of reads and nonzero expressed variables,
3. remove variables based on the mean read counts across samples.

First of all, add metadata for both variables and samples using ASURAT function
`add_metadata()`.

The arguments are

* `sce`: SingleCellExperiment object, and
* `mitochondria_symbol`: a string representing for mitochondrial genes.
```{r}
pbmc <- add_metadata(sce = pbmc, mitochondria_symbol = "^MT-")
```
One can check the results in `rowData(sce)` and `colData(sce)` slots.



<br>

### Remove variables based on expression profiles
ASURAT function `remove_variables()` removes variable (gene) data such that
the numbers of non-zero expressing samples (cells) are less than `min_nsamples`.
```{r}
pbmc <- remove_variables(sce = pbmc, min_nsamples = 10)
```



<br>

### Remove samples based on expression profiles
Qualities of sample (cell) data are confirmed based on proper visualization of
`colData(sce)`.
ASURAT function `plot_dataframe2D()` shows scatter plots of two-dimensional data.
See `?plot_dataframe2D` for details.
```{r}
dataframe2D <- data.frame(x = colData(pbmc)$nReads, y = colData(pbmc)$nGenes)
plot_dataframe2D(dataframe2D = dataframe2D) +
  ggplot2::labs(x = "Number of reads", y = "Number of genes") +
  ggplot2::theme_classic(base_size = 20)
```

```{r}
dataframe2D <- data.frame(x = colData(pbmc)$nReads, y = colData(pbmc)$percMT)
plot_dataframe2D(dataframe2D = dataframe2D) +
  ggplot2::labs(x = "Number of reads", y = "Perc of MT reads") +
  ggplot2::theme_classic(base_size = 20)
```

ASURAT function `remove_samples()` removes sample (cell) data by setting cutoff
values for the metadata.

The arguments are

* `sce`: SingleCellExperiment object,
* `min_nReads` and `max_nReads`: minimum and maximum number of reads,
* `min_nGenes` and `max_nGenes`: minimum and maximum number of non-zero
expressed genes, and
* `min_percMT` and `max_percMT`: minimum and maximum percent of reads that
map to mitochondrial genes, respectively.
If there is no mitochondrial genes, set them as `NULL`.
```{r}
pbmc <- remove_samples(sce = pbmc, min_nReads = 3500, max_nReads = 20000,
                       min_nGenes = 100, max_nGenes = 1e+10,
                       min_percMT = 0, max_percMT = 10)
```

**Tips:**
Take care not to set excessive values for the arguments of `remove_samples()`,
or it may produce biased results.



<br>

### Remove variables based on the mean read counts
Qualities of variable (gene) data are confirmed based on proper visualization of
`rowData(sce)`.
ASURAT function `plot_dataframe2D()` shows scatter plots of two-dimensional data.
See `?plot_dataframe2D` for details.
```{r}
dataframe2D <- data.frame(x = 1:nrow(rowData(pbmc)),
                          y = sort(rowData(pbmc)$nSamples, decreasing = TRUE))
plot_dataframe2D(dataframe2D = dataframe2D) +
  ggplot2::labs(x = "Rank of genes", y = "Mean read counts") +
  ggplot2::theme_classic(base_size = 20)
```

ASURAT function `remove_variables_second()` removes variable (gene) data
such that the mean read counts across samples are less than `min_meannReads`.
```{r}
pbmc <- remove_variables_second(sce = pbmc, min_meannReads = 0.05)
```



<br>

## Normalize data
Normalize data using log transformation with a pseudo count and center the data
by subtracting with the mean expression levels across samples (cells).
The resulting normalized-and-centered data are used for downstream analyses.

Perform log-normalization with a pseudo count.
```{r}
assay(pbmc, "logcounts") <- log(assay(pbmc, "counts") + 1)
```

Center each row data.
```{r}
mat <- assay(pbmc, "logcounts")
assay(pbmc, "centered") <- sweep(mat, 1, apply(mat, 1, mean), FUN = "-")
```



<br>

# Multifaceted sign analysis {#sign}
Infer cell or disease types, biological functions, and signaling pathway
activity at the single-cell level by inputting related databases.

ASURAT transforms centered read count tables to functional feature matrices,
termed sign-by-sample matrices (SSMs).
Using SSMs, perform unsupervised clustering of samples (cells).



<br>

## Load databases
Load databases.
```{r}
urlpath <- "https://github.com/keita-iida/ASURATDB/blob/main/genes2bioterm/"
load(url(paste0(urlpath, "20220108_human_COMSig.rda?raw=TRUE"))) # CO & MSigDB
load(url(paste0(urlpath, "20201213_human_GO_red.rda?raw=TRUE"))) # GO
load(url(paste0(urlpath, "20201213_human_KEGG.rda?raw=TRUE")))   # KEGG
```
The reformatted knowledge-based data were available from the following
repositories:

* [DOI:10.6084/m9.figshare.19102598](https://figshare.com/s/0599d2de970c2deb675c)
* [Github ASURATDB](https://github.com/keita-iida/ASURATDB)

Prepare a correlation matrix of gene expressions using random sampling for the
fast computation.
```{r}
set.seed(1)
nrand_samples <- 1000
mat <- t(as.matrix(assay(pbmc, "centered")))
ids <- sample(rownames(mat), nrand_samples, prob = NULL)
pbmc_cormat <- cor(mat[ids, ], method = "spearman")
```

Set gene expression data into `altExp(sce)`.

**Tips:**
Take care not to use a slot name "log-normalized" for `altExp(sce)`,
which may produce an error when using a Seurat (version 4.0.5) function
`as.Seurat()` in the downstream analysis.
```{r}
sname <- "logcounts"
altExp(pbmc, sname) <- SummarizedExperiment(list(counts = assay(pbmc, sname)))
```

Add ENTREZ Gene IDs to `rowData(sce)`.
```{r, message = FALSE, warning = FALSE}
dictionary <- AnnotationDbi::select(org.Hs.eg.db::org.Hs.eg.db,
                                    key = rownames(pbmc),
                                    columns = "ENTREZID", keytype = "SYMBOL")
dictionary <- dictionary[!duplicated(dictionary$SYMBOL), ]
rowData(pbmc)$geneID <- dictionary$ENTREZID
```

Add formatted databases to `metadata(sce)$sign`.
```{r}
pbmcs <- list(CM = pbmc, GO = pbmc, KG = pbmc)
metadata(pbmcs$CM) <- list(sign = human_COMSig[["cell"]])
metadata(pbmcs$GO) <- list(sign = human_GO[["BP"]])
metadata(pbmcs$KG) <- list(sign = human_KEGG[["pathway"]])
```



<br>

## Create signs
ASURAT function `remove_signs()` redefines functional gene sets for the
input database by removing genes, which are not included in `rownames(sce)`,
and further removes biological terms including too few or too many genes.

The arguments are

* `sce`: SingleCellExperiment object,
* `min_ngenes`: minimal number of genes> 1 (the default value is 2), and
* `max_ngenes`: maximal number of genes> 1 (the default value is 1000).
```{r}
pbmcs$CM <- remove_signs(sce = pbmcs$CM, min_ngenes = 2, max_ngenes = 1000)
pbmcs$GO <- remove_signs(sce = pbmcs$GO, min_ngenes = 2, max_ngenes = 1000)
pbmcs$KG <- remove_signs(sce = pbmcs$KG, min_ngenes = 2, max_ngenes = 1000)
```
The results are stored in `metadata(sce)$sign`.

ASURAT function `cluster_genes()` clusters functional gene sets using
a correlation graph-based decomposition method, which produces strongly,
variably, and weakly correlated gene sets (SCG, VCG, and WCG, respectively).

The arguments are

* `sce`: SingleCellExperiment object,
* `cormat`: correlation matrix of gene expressions,
* `th_posi` and `th_nega`: threshold values of positive and negative
correlation coefficients, respectively.

**Tips:**
Empirically, typical values of `th_posi` and `th_nega` are
$0.15 \le {\rm th{\_}posi} \le 0.4$ and $-0.4 \le {\rm th{\_}nega} \le -0.15$,
but one cannot avoid trial and error for setting these values.
An exhaustive parameter searching is time-consuming but helpful for obtaining
interpretable results.
```{r}
set.seed(1)
pbmcs$CM <- cluster_genesets(sce = pbmcs$CM, cormat = pbmc_cormat,
                             th_posi = 0.30, th_nega = -0.30)
set.seed(1)
pbmcs$GO <- cluster_genesets(sce = pbmcs$GO, cormat = pbmc_cormat,
                             th_posi = 0.30, th_nega = -0.30)
set.seed(1)
pbmcs$KG <- cluster_genesets(sce = pbmcs$KG, cormat = pbmc_cormat,
                             th_posi = 0.20, th_nega = -0.20)
```
The results are stored in `metadata(sce)$sign`.

ASURAT function `create_signs()` creates signs by the following criteria:

1. the number of genes in SCG>= `min_cnt_strg` (the default value is 2) and
2. the number of genes in VCG>= `min_cnt_vari` (the default value is 2),

which are independently applied to SCGs and VCGs, respectively.

**Tips:**
Empirically, typical values of `min_cnt_strg` and `min_cnt_vari` are
$2 \le {\rm min\_cnt\_strg} = {\rm min\_cnt\_vari} \le 3$,
but one cannot avoid trial and error for setting these values.
An exhaustive parameter searching is time-consuming but helpful for obtaining
interpretable results.
```{r}
pbmcs$CM <- create_signs(sce = pbmcs$CM, min_cnt_strg = 2, min_cnt_vari = 2)
pbmcs$GO <- create_signs(sce = pbmcs$GO, min_cnt_strg = 3, min_cnt_vari = 3)
pbmcs$KG <- create_signs(sce = pbmcs$KG, min_cnt_strg = 3, min_cnt_vari = 3)
```
The results are stored in `metadata(sce)$sign_all`,
where "CorrType" indicates SCG or VCG, "Corr" means the average correlation
coefficients of SCG or VCG, "CorrWeak" means the average correlation
coefficients of WCG, "CorrGene" means SCG or VCG, and "WeakCorrGene" means WCG.
The orders of gene symbols and ENTREZ IDs, separated by "/", are consistent.

**Tips:**
If one would like to recreate signs, reset the list of objects by, e.g.,
(`pbmcs <- list(CM = pbmc, GO = pbmc, KG = pbmc)`), and go back to
`remove_signs()`.



<br>

## Select useful signs
If signs have semantic similarity information, one can use ASURAT function
`remove_signs_redundant()` for removing redundant sings using the semantic
similarity matrices.

The arguments are

* `sce`: SingleCellExperiment object,
* `similarity_matrix`: a semantic similarity matrix,
* `threshold`: a threshold value of semantic similarity, used for regarding
biological terms as similar ones, and
* `keep_rareID`: if `TRUE`, biological terms with the larger ICs are kept.

**Tips:**
The optimal value of `threshold` depends on the ontology structure as well as
the method for computing semantic similarity matrix.
```{r}
pbmcs$GO <- remove_signs_redundant(sce = pbmcs$GO,
                                   similarity_matrix = human_GO$similarity_matrix$BP,
                                   threshold = 0.80, keep_rareID = TRUE)
```
The results are stored in `metadata(sce)$sign_SCG`, `metadata(sce)$sign_VCG`,
`metadata(sce)$sign_all`, and if there exist, `metadata(sce)$sign_SCG_redundant`
and `metadata(sce)$sign_VCG_redundant`.

ASURAT function `remove_signs_manually()` removes signs by specifying
IDs (e.g., `GOID:XXX`) or descriptions (e.g., `metabolic`) using `grepl()`.
The arguments are `sce` and `keywords` (keywords separated by `|`).
```{r}
keywords <- "Covid|COVID|foofoo|hogehoge"
pbmcs$KG <- remove_signs_manually(sce = pbmcs$KG, keywords = keywords)
```
The results are stored in `metadata(sce)$sign_SCG`, `metadata(sce)$sign_VCG`,
and `metadata(sce)$sign_all`.

There is another ASURAT function `select_signs_manually()`, a counter part of
`remove_signs_manually()`, which removes signs that do not include `keywords`
(keywords separated by `|`).
```{r, eval = FALSE}
keywords <- "cell|cyte"
test <- select_signs_manually(sce = pbmcs$CM, keywords = keywords)
```
The results are stored in `metadata(sce)$sign_SCG`, `metadata(sce)$sign_VCG`,
and `metadata(sce)$sign_all`.



<br>

## Create sign-by-sample matrices
ASURAT function `create_sce_signmatrix()` creates a new SingleCellExperiment
object `new_sce`, consisting of the following information:

* `assayNames(new_sce)`: counts (SSM whose entries are termed sign scores),
* `names(colData(new_sce))`: nReads, nGenes, percMT,
* `names(rowData(new_sce))`: ParentSignID, Description, CorrGene, etc.,
* `names(metadata(new_sce))`: sign_SCG, sign_VCG, etc.,
* `altExpNames(new_sce)`: something if there is data in `altExp(sce)`.

The arguments are

* `sce`: SingleCellExperiment object,
* `weight_strg`: weight parameter for SCG (the default value is 0.5), and
* `weight_vari`: weight parameter for VCG (the default is 0.5).
```{r}
pbmcs$CM <- makeSignMatrix(sce = pbmcs$CM, weight_strg = 0.5, weight_vari = 0.5)
pbmcs$GO <- makeSignMatrix(sce = pbmcs$GO, weight_strg = 0.5, weight_vari = 0.5)
pbmcs$KG <- makeSignMatrix(sce = pbmcs$KG, weight_strg = 0.5, weight_vari = 0.5)
```

Below are head and tail of `assay(sce, "counts")`:
```{r}
rbind(head(assay(pbmcs$CM, "counts")[, 1:3], n = 4),
      tail(assay(pbmcs$CM, "counts")[, 1:3], n = 4))
```



<br>

## Reduce dimensions of sign-by-sample matrices {#dim_reduction}
Perform principal component analysis and t-distributed stochastic neighbor
embedding.
```{r}
mat <- t(as.matrix(assay(pbmcs$CM, "counts"))) ; set.seed(1)
res <- Rtsne::Rtsne(mat, dim = 2, pca = TRUE, initial_dims = 30)
reducedDim(pbmcs$CM, "TSNE") <- res[["Y"]]

mat <- t(as.matrix(assay(pbmcs$GO, "counts"))) ; set.seed(1)
res <- Rtsne::Rtsne(mat, dim = 2, pca = TRUE, initial_dims = 30)
reducedDim(pbmcs$GO, "TSNE") <- res[["Y"]]

mat <- t(as.matrix(assay(pbmcs$KG, "counts"))) ; set.seed(1)
res <- Rtsne::Rtsne(mat, dim = 3, pca = TRUE, initial_dims = 50)
reducedDim(pbmcs$KG, "TSNE") <- res[["Y"]]
```

The results can be visualized by ASURAT functions `plot_dataframe2D()` or
`plot_dataframe3D()`.
See `?plot_dataframe2D` and `?plot_dataframe3D` for details.
```{r}
dataframe2D <- as.data.frame(reducedDim(pbmcs$CM, "TSNE"))
plot_dataframe2D(dataframe2D = dataframe2D) +
  ggplot2::labs(title = "PBMC (CO MSigDB)", x = "tSNE_1", y = "tSNE_2") +
  ggplot2::theme_classic(base_size = 15)
```

```{r}
dataframe2D <- as.data.frame(reducedDim(pbmcs$GO, "TSNE"))
plot_dataframe2D(dataframe2D = dataframe2D) +
  ggplot2::labs(title = "PBMC (GO)", x = "tSNE_1", y = "tSNE_2") +
  ggplot2::theme_classic(base_size = 15)
```

```{r}
dataframe3D <- as.data.frame(reducedDim(pbmcs$KG, "TSNE"))
plot_dataframe3D(dataframe3D = dataframe3D,
                 theta = 45, phi = 30, title = "PBMC (KEGG)",
                 xlabel = "tSNE_1", ylabel = "tSNE_2", zlabel = "tSNE_3")
```



<br>

## Cluster cells
### Use Seurat functions
To date (December, 2021), one of the most useful clustering methods in scRNA-seq
data analysis is a combination of a community detection algorithm and
graph-based unsupervised clustering, developed in Seurat package.

In this tutorial, our strategy is as follows:

1. convert SingleCellExperiment objects into Seurat objects
  (note that `rowData()` and `colData()` must have data),
2. perform `ScaleData()`, `RunPCA()`, `FindNeighbors()`, and `FindClusters()`,
3. convert Seurat objects into temporal SingleCellExperiment objects `temp`,
4. add `colData(temp)$seurat_clusters` into `colData(sce)$seurat_clusters`.
```{r, message = FALSE, warning = FALSE, results = "hide"}
resolutions <- c(0.20, 0.18, 0.30)
ds <- list(seq_len(30), seq_len(30), seq_len(30))
for(i in seq_len(length(pbmcs))){
  pbmc_surt <- Seurat::as.Seurat(pbmcs[[i]], counts = "counts", data = "counts")
  pbmc_surt[["SSM"]] <- Seurat::CreateAssayObject(
    counts = as.matrix(assay(pbmcs[[i]], "counts")))
  Seurat::DefaultAssay(pbmc_surt) <- "SSM"
  pbmc_surt <- Seurat::ScaleData(pbmc_surt, features = rownames(pbmc_surt))
  pbmc_surt <- Seurat::RunPCA(pbmc_surt, features = rownames(pbmc_surt))
  pbmc_surt <- Seurat::FindNeighbors(pbmc_surt, reduction = "pca", dims = ds[[i]])
  pbmc_surt <- Seurat::FindClusters(pbmc_surt, resolution = resolutions[i])
  pbmc_temp <- Seurat::as.SingleCellExperiment(pbmc_surt)
  colData(pbmcs[[i]])$seurat_clusters <- colData(pbmc_temp)$seurat_clusters
}
```

The results can be visualized by ASURAT functions `plot_dataframe2D()` or
`plot_dataframe3D()`.
See `?plot_dataframe2D` and `?plot_dataframe3D` for details.
```{r}
labels <- colData(pbmcs$CM)$seurat_clusters
dataframe2D <- as.data.frame(reducedDim(pbmcs$CM, "TSNE"))
plot_dataframe2D(dataframe2D = dataframe2D, labels = labels, colors = NULL) +
  ggplot2::labs(title = "PBMC (CO MSigDB)", x = "tSNE_1", y = "tSNE_2", color = "") +
  ggplot2::theme_classic(base_size = 15) +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 4)))
```

```{r}
labels <- colData(pbmcs$GO)$seurat_clusters
dataframe2D <- as.data.frame(reducedDim(pbmcs$GO, "TSNE"))
plot_dataframe2D(dataframe2D = dataframe2D, labels = labels, colors = NULL) +
  ggplot2::labs(title = "PBMC (GO)", x = "tSNE_1", y = "tSNE_2", color = "") +
  ggplot2::theme_classic(base_size = 15) +
  ggplot2::guides(colour = ggplot2::guide_legend(override.aes = list(size = 4)))
```

```{r}
labels <- colData(pbmcs$KG)$seurat_clusters
dataframe3D <- as.data.frame(reducedDim(pbmcs$KG, "TSNE")[, seq_len(3)])
plot_dataframe3D(dataframe3D = dataframe3D, labels = labels, colors = NULL,
                 theta = 45, phi = 30, title = "PBMC (KEGG)",
                 xlabel = "tSNE_1", ylabel = "tSNE_2", zlabel = "tSNE_3")
```



<br>

### Cell cycle inference using Seurat functions
If there is gene expression data in `altExp(sce)`, one can infer cell cycle
phases by using Seurat functions in the similar manner as above.
```{r, message = FALSE, warning = FALSE}
pbmc_surt <- Seurat::as.Seurat(pbmcs$CM, counts = "counts", data = "counts")
pbmc_surt[["GEM"]] <- Seurat::CreateAssayObject(
  counts = as.matrix(assay(altExp(pbmcs$CM), "counts")))
Seurat::DefaultAssay(pbmc_surt) <- "GEM"
pbmc_surt <- Seurat::ScaleData(pbmc_surt, features = rownames(pbmc_surt))
pbmc_surt <- Seurat::RunPCA(pbmc_surt, features = rownames(pbmc_surt))
pbmc_surt <- Seurat::CellCycleScoring(pbmc_surt,
                                      s.features = Seurat::cc.genes$s.genes,
                                      g2m.features = Seurat::cc.genes$g2m.genes)
pbmc_temp <- Seurat::as.SingleCellExperiment(pbmc_surt)
colData(pbmcs$CM)$Phase <- colData(pbmc_temp)$Phase
```



<br>

## Investigate significant signs {#separation_index}
Significant signs are analogous to differentially expressed genes but bear
biological meanings.
Note that naïve usages of statistical tests should be avoided because the row
vectors of SSMs are centered.

Instead, ASURAT function `compute_sepI_all()` computes separation indices
for each cluster against the others.
Briefly, a separation index "sepI", ranging from -1 to 1, is a nonparametric
measure of significance of a given sign score for a given subpopulation.
The larger (resp. smaller) sepI is, the more reliable the sign is as a positive
(resp. negative) marker for the cluster.

The arguments are

* `sce`: SingleCellExperiment object,
* `labels`: a vector of labels of all the samples, and
* `nrand_samples`: an integer for the number of samples used for random
  sampling, which samples at least one sample per cluster.
```{r, message = FALSE, results = "hide"}
for(i in seq_len(length(pbmcs))){
  set.seed(1)
  labels <- colData(pbmcs[[i]])$seurat_clusters
  pbmcs[[i]] <- compute_sepI_all(sce = pbmcs[[i]], labels = labels,
                                 nrand_samples = 100)
}
```
The results are stored in `metadata(sce)$marker_signs`.

When computing separation indices between given clusters, e.g., cluster 1 versus
clusters 2 and 3, use an ASURAT function `compute_sepI_clusters()`.
See `?compute_sepI_clusters` for details.



<br>

## Investigate significant genes
### Use Seurat function
To date (December, 2021), one of the most useful methods of multiple statistical
tests in scRNA-seq data analysis is to use a Seurat function `FindAllMarkers()`.

If there is gene expression data in `altExp(sce)`, one can investigate
differentially expressed genes by using Seurat functions in the similar manner
as described before.
```{r, message = FALSE}
set.seed(1)
pbmc_surt <- Seurat::as.Seurat(pbmcs$CM, counts = "counts", data = "counts")
pbmc_surt[["GEM"]] <- Seurat::CreateAssayObject(
  counts = as.matrix(assay(altExp(pbmcs$CM), "counts")))
Seurat::DefaultAssay(pbmc_surt) <- "GEM"
pbmc_surt <- Seurat::SetIdent(pbmc_surt, value = "seurat_clusters")
res <- Seurat::FindAllMarkers(pbmc_surt, only.pos = TRUE,
                              min.pct = 0.25, logfc.threshold = 0.25)
metadata(pbmcs$CM)$marker_genes$all <- res
```

ASURAT function `plog_violin()` shows violin plots showing gene expression
distributions.
See `?plot_violin` for details.
```{r}
vname <- "CD79A"
expr_sub <- altExp(pbmcs$CM, "logcounts")
expr_sub <- expr_sub[rownames(expr_sub) == vname]
labels <- colData(pbmcs$CM)$seurat_clusters
dataframe1D <- as.data.frame(t(assay(expr_sub, "counts")))
plot_violin(dataframe1D = dataframe1D, labels = labels, colors = NULL) +
  ggplot2::labs(title = vname, x = "Cluster (CO MSigDB)",
                y = "Normalized expression") +
  ggplot2::theme_classic(base_size = 20) +
  ggplot2::theme(legend.position = "none")
```



<br>

## Multifaceted analysis {#multilayered}
Simultaneously analyze multiple sign-by-sample matrices, which helps us
characterize individual samples (cells) from multiple biological aspects.

First, manually curate significant signs by setting keywords, followed by
selecting top significant signs and genes for the clustering results
with respect to separation index and p-value, respectively.
```{r}
# Significant signs
marker_signs <- list()
keywords <- c("LIMB|limb|OENOCYTE|oenocyte", "GO:0002221-S|GO:0035425-S",
              "path:hsa00565-S|path:hsa00564-S")
for(i in seq_len(length(pbmcs))){
  marker_signs[[i]] <- metadata(pbmcs[[i]])$marker_signs$all
  marker_signs[[i]] <- marker_signs[[i]][!grepl(keywords[i],
                                                marker_signs[[i]]$Description), ]
  marker_signs[[i]] <- dplyr::group_by(marker_signs[[i]], Ident_1)
  marker_signs[[i]] <- dplyr::slice_max(marker_signs[[i]], sepI, n = 1)
  marker_signs[[i]] <- dplyr::slice_min(marker_signs[[i]], Rank, n = 1)
}
# Significant genes
marker_genes_CM <- metadata(pbmcs$CM)$marker_genes$all
marker_genes_CM <- dplyr::group_by(marker_genes_CM, cluster)
marker_genes_CM <- dplyr::slice_min(marker_genes_CM, p_val_adj, n = 5)
marker_genes_CM <- dplyr::slice_max(marker_genes_CM, avg_log2FC, n = 1)
```

ASURAT function `plot_multiheatmaps()` shows heatmaps (ComplexHeatmap object)
of sign scores and gene expression levels (if there are), where rows and columns
stand for sign (or gene) and sample (cell), respectively.
See `?plot_multiheatmaps` for details.

The arguments are prepared as follows:
```{r}
# ssm_list
pbmcs_sub <- list() ; ssm_list <- list()
for(i in seq_len(length(pbmcs))){
  pbmcs_sub[[i]] <- pbmcs[[i]][rownames(pbmcs[[i]]) %in% marker_signs[[i]]$SignID, ]
  ssm_list[[i]] <- assay(pbmcs_sub[[i]], "counts")
}
names(ssm_list) <- c("SSM_COMSig", "SSM_GO", "SSM_KEGG")
# gem_list
expr_sub <- altExp(pbmcs$CM, "logcounts")
expr_sub <- expr_sub[rownames(expr_sub) %in% marker_genes_CM$gene]
gem_list <- list(GeneExpr = assay(expr_sub, "counts"))
# ssmlabel_list
labels <- list() ; ssmlabel_list <- list()
for(i in seq_len(length(pbmcs))){
  labels[[i]] <- data.frame(label = colData(pbmcs_sub[[i]])$seurat_clusters)
  labels[[i]]$color <- NA
  ssmlabel_list[[i]] <- labels[[i]]
}
names(ssmlabel_list) <- c("Label_COMSig", "Label_GO", "Label_KEGG")
# gemlabel_list
label_CC <- data.frame(label = colData(pbmcs$CM)$Phase, color = NA)
gemlabel_list <- list(CellCycle = label_CC)
```

Show heatmaps for the selected signs and genes.
```{r, message = FALSE, warning = FALSE}
set.seed(1)
plot_multiheatmaps(ssm_list = ssm_list, gem_list = gem_list,
                   ssmlabel_list = ssmlabel_list, gemlabel_list = gemlabel_list,
                   nrand_samples = 100, show_row_names = TRUE, title = "PBMC")
```

ASURAT function `plot_violin()` shows violin plots of sign score distributions,
which helps users annotate each cluster.
See `?plot_violin` for details.
```{r}
vlist <- list(c("CM", "CL:0000798-S", "gamma-delta T cell"),
              c("CM", "CL:0000623-V", "natural killer cell"),
              c("GO", "GO:0006910-S", "phagocytosis, recognition"),
              c("CM", "MSigID:12-S", "AIZARANI...B_CELLS..."),
              c("CM", "CL:0000912-S", "helper T cell"),
              c("KG", "path:hsa04666-V", "Fc gamma R-mediated phagocytosis"))
xlabels <- list(CM = "Cluster (CO MSigDB)", GO = "Cluster (GO)", KG = "Cluster (KEGG)")
p <- list()
for(i in seq_len(length(vlist))){
  ind <- which(rownames(pbmcs[[vlist[[i]][1]]]) == vlist[[i]][2])
  pbmc_sub <- pbmcs[[vlist[[i]][1]]][ind, ]
  labels <- colData(pbmc_sub)$seurat_clusters
  dataframe1D <- as.data.frame(t(assay(pbmc_sub, "counts")))
  p[[i]] <- plot_violin(dataframe1D = dataframe1D, labels = labels) +
    ggplot2::labs(title = paste0(vlist[[i]][2], "\n", vlist[[i]][3]),
                  x = xlabels[[vlist[[i]][1]]], y = "Sign score", fill = "Cluster") +
    ggplot2::theme(legend.position = "none")
}
gridExtra::grid.arrange(p[[1]], p[[2]], p[[3]], p[[4]], p[[5]], p[[6]])
```



<br>

# Session information

```{r}
sessionInfo()
```
